Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*

### 2022-09-13 paper/paper.qmd
edits

---
title: Clinical Phenotype and Survival Analysis in Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*
author:
  - name: Dean Marchiori
    email: deanmarchiori@gmail.com
    attributes:
        corresponding: true
abstract: |
  A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding of associations present in the heterogenous features of the clinical phenotype, and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data on reported cases. This research aims to collate and standardise available case reports into a common dataset, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To account for censoring in the data Kaplan-Meier estimation is formulated calculate patient survival time. The analysis correctly detected the classic phenotype for this condition, as well as an additional distinct feature-cluster relating to findings of vocal cord paralysis, feeding dysfunction and respiratory failure. Several additional outlying and isolated clinical features were also identified. The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years. This wide range of outcomes is provisionally associated with different variant types however this conclusion could not be validated based on very low sample sizes. This analysis represents the first of its kind to help describe associations and trajectories of individuals with this newly reported condition, despite challenges with sparse and inconsistent data. This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.
keywords: 
  - NEDMABA
  - SMPD4  
#journal: "Preprint"    
date: last-modified
execute: 
  echo: false 
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

@ravenscroft2021neurogenetic performed a study over 190 probands and identified a novel missense variation in SMPD4 which at the time of whole exome sequencing was not well described in the literature. The impacted family from Melbourne exhibited features involving arthrogryposis multiplex congenita, complex brain malformations, small for gestation age and hypoplasia of the corpus callosum. In two of the three related cases were additional features of microcephaly, congenital encephalopathy, cerebellar malformation and hypoplasia and hypomyelination.

A detailed study by @magini2019loss involved 12 unrelated families with 32 individuals (21 with detailed clinical information). Presentations were of microcephaly, simplified gyral pattern of the cortex, hypomyelination, cerebellar hypoplasia, congenital arthrogryposis, and early fetal/postnatal demise. Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly. Three missense changes were noted in the study with affected children in these families often showing a milder presentation suggestive of possible residual function. In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI. Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment.

A case study from China is described in @ji2022case involving a girl presenting in infancy with intrauterine growth restriction, microcephaly, postnatal developmental delay, arthrogryposis, hypertonicity, seizure, and hypomyelination on brain magnetic resonance imaging. The authors report parallels with two cases showing the same homozygous null variant. An individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features.

A recent study by @bijarnia2022growth presents the case of a 22-month old girl presenting with the typical phenotype of neurodevelopmental delay, prenatal onset growth failure, arthrogryposis, microcephaly and brain anomalies including severe hypomyelination, simplified gyral pattern and hypoplasia of corpus callosum and brainstem. Notably, there was also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.

Further work to collate and analyse data relating to NEDMABA is challenging. While @magini2019loss cataloged a detailed clinical phenotype data set as a supplementary data set to their study, these data are not suitable for statistical analysis as presented. Many of the clinical features are represented as free text descriptions and a variety of non-standard terminologies are applied between cases, making machine interpretation of the data difficult. Other studies [@ji2022case; @bijarnia2022growth] present only written case reports with some tabulated summaries. While larger studies [@ravenscroft2021neurogenetic; @monies2019lessons] focus more on genetic data, with less detail included on the clinical presentation of individuals in the cohort.

To overcome these data challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell. In this case detailed text-based data can be transformed to high dimensional binary indicators of key clinical features. This structure is conducive to effective statistical analysis and integrates deliberately with the *tidyverse* [@tidyverse] collection of packages for data analysis in R [@rbase].

Further analysis to better describe rare genetic conditions was explored in @diaz2020phenotype in the context of large scale genotype-phenotype analysis. The authors argue patients with rare disorders (often in small samples) frequently present with varied symptoms that do not match exactly with the described phenotype. When considering this situation of low sample size data with many features, methods in the field of multivariate statistical analysis are commonly deployed. Here the aim is to find substructure in the data, or a simple representation of the multi-dimensional space. Methods, such as Multiple Correspondence Analysis [@le2010multiple] are appropriate for high dimensional data comprised of binary indicators. This technique is cited in many studies in the application of dimension reduction and clustering of comorbidities and phenotypes from disease [@han2018cluster; @costa2013use].

@@

An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in *SMPD4*. With current research highlighting a diverse and heterogeneous phenotype, the relationship and correlations between these diverse features is not well understood. Furthermore, the survival time of affected children has not been extensively analysed despite early-demise featuring as a severe outcome in many cases. While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.

This research has three key aims. First, to collate and transform the variety of early case reports and studies on clinical phenotype data for this novel variant into an analysis-ready *tidy* data set. Secondly, to conduct analysis on the associations between commonly reported clinical features. Finally, to statistically quantify the expected survival time of children with NEDMABA based on the current case reports.

@@

The data were read into `R` [@rbase] without any changes, so as to preserve the reproducibility of the data transformation steps. The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item). This resulted in a long-formatted data set of 21 rows (individuals) and 64 columns (clinical features). In many cases, key clinical information was entered as free-text descriptions, which rendered any attempt of meaningful analysis impractical (@tbl-nontidy). In these cases, the text was tokenised by separating the list of clinical observations at each comma and forming a binary indicator column noting its presence '1' or absence '0' (@tbl-tidy).

@@

The final data set consisted of 28 observations (one per individual) and 152 variables (one per clinical feature). These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various bio markers. A full list of this data set is available in @tbl-dict. These data sets were compiled into a publicly available `R` Package called `SMPD4` [@smpd4-data] in order to allow for reproducibility and sharing.

@@

The data on subjects introduced above is subset to include only variables that can be transformed into binary indicator variables that represent the presence or absence of a given clinical feature. Only those features that appeared in more than one case report were included to minimise the influence of isolated features in this analysis.

@@

```{r ctr-plot, fig.height=6, fig.width = 8, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  mutate(vars = str_remove(vars, "_[:digit:]")) |> 
  mutate(vars = str_remove(vars, "_ind")) |> 
  mutate(vars = str_replace_all(vars, "_", " ")) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered()+
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  theme(legend.position = 'bottom') +
  labs(y = '')
```

@@

Cluster validation measures resulted in either 2 or 5 clusters as the optimal choice (@tbl-opticlust). In order to partition the data into meaningful groupings, k-means with 5 clusters was selected. This reflects the optimal choice for both AD and FOM cluster metrics. A scatter plot of the cluster results shown in @fig-clust is projected on the first two principal dimensions highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies. A separate and adjacent cluster is formed of airway and feeding related conditions involving vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles. Three other smaller satellite clusters exists where highly distinctive features were associated with a single individual or family. A full articulation of feature to cluster assignment is given in @tbl-profile.

```{r}
#| label: tbl-opticlust
#| tbl-cap: "Optimal clusters"
optimalScores(cluster_select) |> 
  as_tibble(rownames = 'Metric') |> 
  gt() |> 
  fmt_number(columns = Score, decimals = 2)
```

@@

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days. At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98). Survival probability after one year at 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).

@@

```{r,fig.width=7, warning=FALSE}
# KM by variant

@@

Often, data collected in studies is fit for descriptive purposes but not for in-depth statistical analysis. This meta-review of current studies was able to consolidate and make publicly available, data sets that allow analysis to be conducted on emerging conditions such as NEDMABA.

The use of dimension reduction techniques were shown to be successful in uncovering relevant and novel substructure in the data. This was particularly evident with the detection of the accepted typical phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay.

After excluding two outlying remarkable cases, this analysis clearly identifies the presence of an adjacent cluster of related features that are not as well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo. We identify the prolific feature of 'respiratory distress' as an expression more commonly associated with *other* features such as vocal cord palsy, tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD). These features are present in the case reported data, but not well described in HPO and help account for the variation seen in the clinical phenotype in the first few principal dimensions.

Other smaller clusters identify outlying features that are expressed in individuals or specific families. Many of the outlying features are highly specific facial or anatomical dysmorphisms. While coarse facial features are a noted and core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail, which skewed the interpretation.

On a similar note, talipes appears as an isolated feature without the typical microcephaly and hypomyelination. However, there was significant heterogeneity in how feet contractures were reported in clinical case reports. While distinctions were made during data pre-processing between rocker-bottom feed and talipes, this likely to make intpretation somewhat difficult. This is particularly clear as rocker-bottom feet was found closely related in our analysis to other core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature.

Another feature clusted within core identified features is bilateral cleft lip. This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features.

The heterogeneous variation in the data was clearly evident however, with the first four principal axes being required to explain more than 50% of the variation in the data set. Visual inspection of the data projected into this lower dimensional space indicate the primary sources of variation in the clinical phenotype are explained by a few individuals with very different clinical features. This has confounded the typical or classical presentation. Not all features in the data were included in this analysis, particular if isolated to just one case. In this circumstance, these terms in @tbl-hpo have been labelled as 'Isolated'.

While the number of clusters chosen all represent valid partition of the data, some metrics indicated an optimal selection of 2 clusters. This type of partition is useful when segregating outliers from the rest of the data and is valid in some applications. In this study, a larger number (5) was selected to align the goals of the analysis which was to uncover distinct groups of features.

In terms of patient survival, the first few months show a sharp decline in survival probability confirming early demise in the infant period as a key factor. A small number of cases show much longer survival into childhood and in some cases into the second decade of life. This work identifies a one year survival rate of 0.403 (95% CI: 0.252, 0.644). On the other hand, a naive estimate using just the number of deceased patients at this time as proportion of the study population is 0.57 (16/28). This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability. The current literature suggests the type of gene variant has an impact on survival time, which seems to be supported in this work. Through a comparison of multiple survival curves, stratified by variation type, the limited sample size prevented definitive conclusions with many cells having expected counts less than 5.

>>>>

---
title: Clinical Phenotype and Survival Analysis of Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies Due to Bi-allelic Loss of Function Variants in *SMPD4*
author:
  - name: Dean Marchiori
    email: deanmarchiori@gmail.com
    attributes:
        corresponding: true
abstract: |
  A recently described, rare genetic condition known as Neurodevelopmental Disorder with Microcephaly, Arthrogryposis, and Structural Brain Anomalies (NEDMABA) has been identified in children with bi-allelic loss-of-function variants in *SMPD4*. The progression of this condition is not well understood with the limited case reports described so far exhibiting a severe and clinically diverse phenotype. A gap exists in the understanding of associations present in the heterogenous features of the clinical phenotype, and the expected survival probabilities of affected individuals. This is driven in part to the paucity of analysis-ready data on reported cases. This analysis aims to collate and standardise available case reports into a common dataset, to analyse and identify meaningful clusters in the clinical phenotype, and to quantify the survival probability for children with NEDMABA. To overcome the challenge of multidimensional data on very few subjects, we employ Multiple Correspondence Analysis (MCA) as a dimension reduction technique, which is then subject to cluster analysis and interpretation. To account for censoring in the data, Kaplan-Meier estimation is formulated to calculate patient survival time. The analysis correctly detected the classic phenotype for this condition, as well as a new distinct feature-cluster relating to findings of vocal cord paralysis, feeding dysfunction and respiratory failure. The survival probability for those affected was found to decline sharply in early infancy with median survival of 150 days, but with some surviving as long as 12.5 years. This wide range of outcomes is provisionally associated with different variant types however this conclusion could not be validated based on very low sample sizes. An `R` package called `SMPD4` was developed to publish standardised analysis-ready datasets used in this study. This analysis represents the first of its kind to help describe associations and trajectories of individuals with this newly reported condition, despite challenges with sparse and inconsistent data. This analysis can provide clinicians and genetic counsellors with better information to aide in decision making and support for families with this rare condition.
keywords: 
  - NEDMABA
  - SMPD4  
#journal: "Preprint"    
date: last-modified
execute: 
  echo: false 
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      formatting: preprint
      model: 3p
      cite-style: authoryear
---

@@

A detailed study by @magini2019loss involved 12 unrelated families with 32 individuals (21 with detailed clinical information). Presentations were of microcephaly, simplified gyral pattern of the cortex, hypomyelination, cerebellar hypoplasia, congenital arthrogryposis, and early fetal/postnatal demise. Despite this being the largest cohort studied, the clinical features and survival times among participants varied greatly. Three missense changes were noted in the study with affected children in these families often showing a milder presentation suggestive of possible residual function. In these cases individuals were able to develop independent motor skills, have mild intellectual disability and arthrogryposis without evidence of simplified gyral patterns on brain MRI. Other patients with truncating variants are shown to have more severe presentations, while a range of additional significant clinical features were reported involving dysmorphic facial features, seizure, vocal cord paralysis and hearing impairment. @ravenscroft2021neurogenetic studied a family from Melbourne exhibiting features involving arthrogryposis multiplex congenita, complex brain malformations, small for gestation age and hypoplasia of the corpus callosum. In two of the three related cases were additional features of microcephaly, congenital encephalopathy, cerebellar malformation and hypoplasia and hypomyelination. A case study from China is described in @ji2022case involving a girl presenting in infancy with intrauterine growth restriction, microcephaly, postnatal developmental delay, arthrogryposis, hypertonicity, seizure, and hypomyelination on brain magnetic resonance imaging. The authors report parallels with two cases showing the same homozygous null variant. An individual reported in @monies2019lessons presented with distinct symptoms of brain atrophy and skeletal dysplasia whereas the case in @magini2019loss with the same variant exhibited more typical clinical features. A recent study by @bijarnia2022growth presents the case of a 22-month old girl presenting with the typical phenotype of neurodevelopmental delay, prenatal onset growth failure, arthrogryposis, microcephaly and brain anomalies including severe hypomyelination, simplified gyral pattern and hypoplasia of corpus callosum and brainstem. Notably, there was also additional non-typical clinical findings of nystagmus and visual impairment secondary to macular dystrophy and retinal pigment epithelial stippling at posterior pole.

Further work to collate and analyse data relating to NEDMABA is challenging. While @magini2019loss included a supplementary clinical phenotype data set in their study, these data are not suitable for statistical analysis. Many of the clinical features are represented as text descriptions and a variety of non-standard terminologies are applied between cases, making machine interpretation of the data difficult. Other studies [@ji2022case; @bijarnia2022growth] present only written case reports with some tabulated summaries. While larger studies [@ravenscroft2021neurogenetic; @monies2019lessons] focus more on genetic data, with less detail included on the clinical presentation of individuals. To overcome these data challenges, @JSSv059i10 proposes a data structure know as 'tidy' data which organises each observation in a row, with each feature as its own column and each value in just one cell. In this case detailed text-based data can be transformed to high dimensional binary indicators of key clinical features. This structure is conducive to effective statistical analysis and integrates deliberately with the *tidyverse* [@tidyverse] collection of packages for data analysis in R [@rbase].

Further analysis to better describe rare genetic conditions was explored in @diaz2020phenotype in the context of large scale genotype-phenotype analysis. The authors argue patients with rare disorders (often in small samples) frequently present with varied symptoms that do not match exactly with the described phenotype. When considering the situation of low sample size data with many features, methods in the field of multivariate statistical analysis are commonly deployed. Here the aim is to find substructure in the data, or a simple representation of the multi-dimensional space. Methods, such as Multiple Correspondence Analysis [@le2010multiple] are appropriate for high dimensional data comprised of binary indicators. This technique is cited in many studies in the application of dimension reduction and clustering of comorbidities and phenotypes from disease [@han2018cluster; @costa2013use].

@@

An open research question exists around the clinical pathway and survival time of children exhibiting bi-allelic loss-of-function variants in *SMPD4*. With current research highlighting a diverse and heterogeneous phenotype, the relationship between these diverse features is not well understood. Furthermore, the survival time of affected children has not been extensively analysed despite early-demise featuring as a severe outcome. While a connection has been highlighted between some missense variants and a milder presentation with longer survival [@magini2019loss], this hypothesis has not been analysed further.

This analysis has three key aims. First, to collate and transform the variety of early case reports and studies on clinical phenotype data for this novel variant into an analysis-ready *tidy* data set. Secondly, to conduct analysis on the associations between commonly reported clinical features. Finally, to statistically quantify the expected survival time of children with NEDMABA based on the current case reports.

@@

The data were read into `R` [@rbase] without any changes, so as to preserve the reproducibility of the data transformation steps. The data were transposed to ensure it could be presented as *tidy* formatted data [@JSSv059i10]. This requires that each variable (clinical phenotype) forms a column; each observation (individual) forms a row and each type of observational unit forms a table (every cell has just one item). This resulted in a long-formatted data set of 21 rows (individuals) and 64 columns (clinical features). In many cases, key clinical information was entered as free-text descriptions, which rendered any attempt of meaningful analysis impractical (@tbl-nontidy). In these cases, the text was tokenised by separating the list of clinical observations at each comma and forming a binary indicator column noting its presence '1' or absence '0' (@tbl-tidy).

@@

The final data set consisted of 28 observations (one per individual) and 152 variables (one per clinical feature). These variables were comprised of qualitative text descriptions, binary indicators of features as described above and other fields including background information on the patient and various bio markers. A full list of this data set is available in @tbl-dict. These data sets were compiled into a publicly available `R` Package called `SMPD4` [@smpd4-data] to allow for reproducibility and sharing.

@@

The data on subjects above is subset to include only variables that can be transformed into binary indicator variables representing the presence or absence of a given clinical feature. Only those features that appeared in more than one case report were included to minimise the influence of isolated features.

@@

```{r ctr-plot, fig.height=6, fig.width = 8, warning=FALSE}
#| label: fig-ctr
#| fig-cap: "Top 10 feature categories from MCA analysis for the first four MCA dimensions. The red dashed line indicates the mean contribution value. Each feature may be present or absent as indicated by 1 or 0 respectively"
mca$var$contrib |> 
  as_tibble(rownames = 'vars') |> 
  select(1:5) |> 
  pivot_longer(cols = -vars, names_to = "dim", values_to = "contribution") |> 
  mutate(col = ifelse(str_detect(vars, "_0"), "absent", "present")) |> 
  group_by(dim) |> 
  top_n(10, wt = contribution) |> 
  mutate(vars = fct_reorder(vars, contribution)) |> 
  mutate(vars = str_remove(vars, "_[:digit:]")) |> 
  mutate(vars = str_remove(vars, "_ind")) |> 
  mutate(vars = str_replace_all(vars, "_", " ")) |> 
  ggplot(aes(contribution, reorder_within(vars, by = contribution, within = dim), fill = col)) +
  geom_col() +
  scale_y_reordered() +
  scale_fill_grey(start = 0.8, end = 0.2) +
  facet_wrap(~dim, scales = "free", ncol = 2) +
  geom_vline(xintercept = 100/122, lty = 2, col = 'red') +
  guides(fill=guide_legend(title="Condition")) +
  theme(legend.position = 'bottom') +
  labs(y = '')
```

@@

Cluster validation measures resulted in either 2 or 5 clusters as the optimal choice (@tbl-opticlust). In order to partition the data into meaningful groupings, k-means with 5 clusters was selected. This reflects the optimal choice for both AD and FOM cluster metrics. The results are summarised in @tbl-clusttab and a scatter plot of the cluster results is shown in @fig-clust is projected on the first two principal dimensions. This highlights a core cluster of classic features of NEDMABA such as microcephaly, hypomyelination, arthrogryposis and feet deformity and structural brain anomalies. A separate and adjacent cluster is formed of airway and feeding related conditions involving vocal cord paralysis, tracheostomy, tube feeding and enlarged ventricles. Three other smaller satellite clusters exists where highly distinctive features were associated with a single individual or family. A full articulation of feature to cluster assignment is given in @tbl-profile.

```{r}
#| label: tbl-opticlust
#| tbl-cap: "Optimal cluster numbers and methods based on a range of stability validation metrics."
optimalScores(cluster_select) |> 
  as_tibble(rownames = 'Metric') |> 
  gt() |> 
  fmt_number(columns = Score, decimals = 2)
```

@@

```{r}
#| label: tbl-clusttab
#| tbl-cap: "A summary table to cluster membership showing the cluster number and number of features in each cluster and a sample of features that have membership in each."
cluster_assignments |> 
  as_tibble(rownames = 'feature') |> 
  mutate(feature = str_remove(feature, "_ind_1|_1"),
         feature = str_replace_all(feature, "_", " ")) |> 
  mutate('Cluster No.' = value) |> 
    mutate("Cluster Description" = fct_recode(as.character(value),
    Core = "4",
    Adjacent = "2",
    Outlying = "3",
    Outlying = "5",
    Outlying = "1",
    )) |> 
  group_by(`Cluster No.`, `Cluster Description`) |> 
  summarise('No. Features' = n(),
            `Sample Features in Cluster` = paste(head(feature, 4), collapse = ", "), .groups = 'drop') |> 
  select(-`Cluster Description`) |> 
  gt() |> 
  gt::tab_footnote(footnote = "A sample of 4 features shown to illustrate cluster membership", locations = cells_column_labels("Sample Features in Cluster")) 
```

@@

The Kaplan-Meier estimator shows a range of survival times (0, 4562 days) with median survival time calculated as 150 days. At time 0 the survival probability was 0.82 (95% CI: 0.69, 0.98). Survival probability after one year of 0.403 (95% CI: 0.252, 0.644) and survival probability at close to ten years of 0.24 (95% CI: 0.11, 0.52).

@@

```{r,fig.width=7, warning=FALSE}
#| label: fig-box
#| fig-cap: "A compariative box plot showing the range of survival times for subjects split by the type of variant. Green bars indicate homozygous variants, while red indicates compound heterozygous variants."

@@

This meta-review of current research was able to consolidate data that allowed statistical analysis to be conducted on affected individuals. The use of dimension reduction techniques was shown to be successful in uncovering relevant and novel substructure in the data. This was particularly evident with the detection of the accepted typical phenotype of microcephaly, hypomyelination, arthrogryposis, simplified gyral pattern and developmental delay. After excluding two outlying remarkable cases, we identified the presence of an adjacent cluster of related features that are not well described in The Human Phenotype Ontology [@kohler2021human] as shown in @tbl-hpo. It shows the prolific feature of 'respiratory distress' as more commonly associated with these adjacent features such as vocal cord palsy, tracheostomy, tube feeding and Gastroesophageal reflux disease (GERD). These features are present in the case reported data, but not well described in HPO and help account for the variation seen in the clinical phenotype in the first few principal dimensions.

The heterogeneous variation in the data was clearly evident with the first four principal axes being required to explain more than 50% of the variation in the data set. Visual inspection shows the primary sources of variation in the clinical phenotype are explained by a few individuals with very different clinical features. Many of these exhibit highly specific facial or anatomical dysmorphisms. While coarse facial features are a core presentation, those with a milder course had clinical reports that articulated these dysmorphisms in much more detail. Additionally, talipes appears as an isolated feature without the typical microcephaly and hypomyelination, however there was significant heterogeneity in how feet contractures were reported in clinical case reports. This is particularly evident in rocker-bottom feet which was found closely related in our analysis to core features despite the prevalence in HPO being quite low (2/21), suggesting possible under-reporting or confounded reporting of this feature. Another core identified feature is bilateral cleft lip. This condition has a reported frequency of just 2/21 (@tbl-hpo) and may be an emerging feature for a more typical presentation given its close relationship with other classic features. Features isolated to just one or few cases in our consolidated data have been labelled as 'Isolated' in @tbl-hpo. An optimal choice of both 2 and 5 clusters were recommended in the analysis. A partition of the data with just two clusters would likely be useful in separating outliers from other data. In this analysis a larger number (5) was selected to align the goals of the analysis which was to uncover distinct groups of features.

In terms of patient survival, the first few months of life show a sharp decline in survival probability confirming early demise in the infant period as a key factor. A small number of cases show survival into childhood and in some cases into the second decade of life. This work identifies a one year survival rate of 0.403 (95% CI: 0.252, 0.644). On the other hand, a naive estimate using just the number of deceased patients at this time as proportion of the study population is 0.43 (1 - 16/28). This highlights the importance of applying a survival analysis technique that incorporates censoring so as not to erroneously overestimate the survival probability. The current literature suggests the type of gene variant has an impact on survival time, which seems to be supported in this work. Through a comparison of multiple survival curves, stratified by variation type, the limited sample size prevented definitive conclusions with many cells having expected counts less than 5.
